Merge biotime timeseries with trait data

### 2021-09-16 analysis/01_filter_by_traits.Rmd
Clean up, annotate

---
title: "Check Coverage"
author: "Kari Norman"
date: "5/20/2019"
output: html_document
---

@@

@@

#Checking Coverage of ID's and trait data -> biotime

check if all bird & mammal biotime data w/id's has an elton match. There are 18 species without a match.
```{r}
no_trait <- biotime_data %>%
  filter(taxa %in% c("Birds", "Mammals")) %>%
  select(id, sourceName, taxa) %>%
  drop_na(id) %>%
  distinct() %>%
  filter(!id %in% elton_bird$id, !id %in% elton_mamm$id)

@@

and 25268 species without an ID, some of which are species codes
```{r}
biotime_data %>%
  #filter(taxa == "Birds") %>%
  select(id, sourceName, taxa) %>%
  filter(is.na(id)) %>%
  distinct() %>%
  dim()
```

Are there any studies for which we have trait data for all species?
```{r}
merge_data <- function(trait_data){
  biotime_data %>%
    select(-scientificName) %>%
    rename(biotimeName = sourceName) %>%
  left_join(trait_data %>% rename(traitName = sourceName), by = "id",
            na_matches = "never") %>%
   drop_na(scientificName)
}
```

Get data w/traits for all studies that have >85% trait coverage

`id` is `NA` if there was no `id` match
`biotimeName` or `traitName` is `NA` if that ID is the result of an average of multliple species matches
`scientificName` is `NA` when there was no trait data for a given ID
```{r}
#set coverage threshold for analysis dataset
percent <- 0.75

@@

trait_merge <- map_dfr(traits, merge_data)

comp_biotime_traits <- trait_merge %>%
  bind_rows(biotime_data %>% 
              filter(!obsID %in% trait_merge$obsID) %>%
              select(-scientificName) %>%
              rename(biotimeName = sourceName))# %>%
  #distinct()

@@

coverage <- comp_biotime_traits %>%
  select(study_id, id, biotimeName, traitName) %>%
  distinct() %>%
  group_by(study_id) %>%
  summarise(perc = sum(!is.na(traitName))/n()) %>%
  arrange(desc(perc))

inc_coverage <- coverage %>% filter(perc > percent)

biotime_traits <- comp_biotime_traits %>%
  #get studies with appropriate coverage and drop observations that we don't have trait data for, and unlabeled non-id'd species
  filter(study_id %in% inc_coverage$study_id, !is.na(scientificName)) %>% 
  group_by(rarefyid, year) %>%
  filter(n_distinct(id) > 4) %>%
  ungroup()
```

Get biotime data filtered by trait coverage
```{r}
bt_cols <- colnames(biotime_data)[!colnames(biotime_data) %in% 
                                         c("sourceName", "obsID", "scientificName")]
biotime_traits <- biotime_traits %>% 
  select(all_of(bt_cols)) %>%
  rename(species = id)
```

Split timeseries that have more than one class of species (split up birds and mammals)
```{r}
#get the class of each species
tax_info <- taxadb::filter_id(unique(biotime_traits$species), "itis")
biotime_traits <- biotime_traits %>% 
  left_join(tax_info %>% select(input, class), by = c("species" = "input"))

@@

#update that studies that needed to be split with new rarefyid breakdowns
bt_traitfiltered <- biotime_traits %>%
  filter(!study_id %in% mix_study) %>%
  bind_rows(mix_split) %>%
  select(-class)
```

```{r}
#save locally and in cache
usethis::use_data(bt_traitfiltered)
pins::pin(bt_traitfiltered, board = "github")
```

>>>>

---
title: "Merge biotime timeseries with trait data"
author: "Kari Norman"
date: "5/20/2019"
output: html_document
---

@@

# Description
This script merges trait data and biotime timeseries data, then filters out timeseries that don't have high enough trait coverage.

@@

check if all bird & mammal biotime data w/id's has an elton match. There are 13 species without a match.
```{r}
no_trait <- biotime_data %>%
  filter(taxa %in% c("Birds", "Mammals")) %>%
  select(id, sourceName, taxa) %>%
  drop_na(id) %>%
  distinct() %>%
  filter(!id %in% elton_bird$id, !id %in% elton_mamm$id)

@@

Function for joining timeseries data with functional trait data. 
```{r}
merge_data <- function(trait_data){
  biotime_data %>%
    select(-scientificName) %>% #remove scientific name from biotime data
    rename(biotimeName = sourceName) %>% #give original biotime name a better column name
  left_join(trait_data %>% rename(traitName = sourceName), by = "id",
            na_matches = "never") %>% # join to trait data
   drop_na(scientificName)
}
```

Get data w/traits for all studies that have >75% trait coverage

`id` is `NA` if there was no `id` match
`biotimeName` or `traitName` is `NA` if that ID is the result of an average of multiple species matches
`scientificName` is `NA` when there was no trait data for a given ID
```{r}
#set coverage threshold for analysis dataset
percent <- 0.75

@@

#map across list of trait sources to merge timeseries with each one
trait_merge <- map_dfr(traits, merge_data)

#combine observations w/trait data back with observations with no trait data
comp_biotime_traits <- trait_merge %>%
  bind_rows(biotime_data %>% 
              filter(!obsID %in% trait_merge$obsID) %>%
              select(-scientificName) %>%
              rename(biotimeName = sourceName))# %>%
  #distinct()

@@

#get a percentage of trait coverage for each study
coverage <- comp_biotime_traits %>%
  select(study_id, id, biotimeName, traitName) %>%
  distinct() %>%
  group_by(study_id) %>%
  summarise(perc = sum(!is.na(traitName))/n()) %>%
  arrange(desc(perc))

#studies to include, since they meet the coverage requirement 
inc_coverage <- coverage %>% filter(perc > percent)

biotime_traits <- comp_biotime_traits %>%
  #get studies with appropriate coverage and drop observations that we don't have trait data for, and unlabeled non-id'd species
  filter(study_id %in% inc_coverage$study_id, !is.na(scientificName)) %>% 
  group_by(rarefyid, year) %>%
  filter(n_distinct(id) > 4) %>% #more than four species for each observation
  ungroup()
```

Get final columns 
```{r}
bt_cols <- colnames(biotime_data)[!colnames(biotime_data) %in% 
                                         c("sourceName", "obsID", "scientificName")]
biotime_traits <- biotime_traits %>% 
  select(all_of(bt_cols)) %>%
  rename(species = id)
```

Some timeseries have observations from more than one class of species (e.g. birds and mammals), here we split them into two individual timeseries.
```{r}
#get the class of each species
tax_info <- taxadb::filter_id(unique(biotime_traits$species), "itis")
biotime_traits <- biotime_traits %>% 
  left_join(tax_info %>% select(input, class), by = c("species" = "input"))

@@

#update the studies that needed to be split with new rarefyid breakdowns
bt_traitfiltered <- biotime_traits %>%
  filter(!study_id %in% mix_study) %>%
  bind_rows(mix_split) %>%
  select(-class)
```

Final merged data set to save
```{r}
#save locally and in cache
usethis::use_data(bt_traitfiltered)
pins::pin(bt_traitfiltered, board = "github")
```
